Supporting Information:\\ NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-08-09 Manuscript/manuscriptPS.tex
Updates in the text

Previous experimental studies have concluded that
PS headgroups are more rigid than phophocholines (PC)
due to the hydrogen bonding network or
electrostatic interactions \cite{browning80,buldt81}.
Multivalent cations and Li$^+$ are able to form strong
dehydrated molecular complexes with PS lipids,
while monovalent ions interact more weakly with PS
containing bilayers \cite{hauser77,kurland79,eisenberg79,hauser83,dluhy83,hauser85,feigenson86,mattai89,roux90,roux91,boettcher11}.
On the other hand, some studies propose that the specific binding
affinity is similar to the negatively charged and zwitterionic lipids and that
the increased cation binding to negatively charged lipid bilayer arise only due
to the increase of local cation concentration in the vicinity of membranes \cite{seelig90,sinn06}.
Dilution of bilayers with PC lipids makes PS headgroups
less rigid and reduces propensity for the formation of
strong complexes with multivalent ions \cite{browning80,buldt81,roux90,roux91}.
The molecular level interpretation of these observations is,
however, not available.

@@

}\label{IONsystems}

@@

}\label{IONsystems}

@@

\subsection{Molecular dynamics simulations}
Molecular dynamics simulation data was collected using
the Open Collaboration method \cite{botan15}.
The NMRlipids project blog (\url{nmrlipids.blogspot.fi}) and
the GitHub repository (\url{github.com/NMRLipids/NMRlipidsIVotherHGs})
were used as the communication platforms.
The simulated systems are listed in Table \ref{IONsystems} and
simulation details are given in the SI.
The simulation data is also indexed in the
searchable database (\url{nmrlipids.fi}),
and in the NMRlipids/MATCH GitHub repository (\url{https://github.com/NMRLipids/MATCH}).

@@

The order parameters of $\alpha$ and $\beta$ carbons in PC lipids
can be used to measure the ion binding affinity because they 
decrease proportionally to the amound of bound positive
charge to a bilayer \cite{akutsu81,altenbach84,seelig87}.
This molecular electrometer concept is especially useful for 
the comparison between simulations and experiments because
the headgroup order parameters can be directly calculated from 
simulations~\cite{catte16}. Also the headgroup order parameters
of negatively charged PS and PG lipids exhibit systemic, but less
characterized dependence on the bound charge~\cite{borle85,macdonald87,roux86,roux90}.
Therefore, the ion binding affinity to negatively charged bilayers
can be better characterized measuring the PC headgroup order parameters from 
mixed bilayers, see section \ref{electrometerFORmixtures} in the supplementary information.

@@

\subsection{Headgroup and glycerol backbone order parameters POPS from $^{13}$C NMR}

The INEPT and 2D R-PDLF experiments from POPS sample give well resolved spectras for all the
carbons in headgroup and glycerol backbone region, except for g$_3$ for which the resolution
was not sufficient to determine the numerical value of the order paramater (Fig. \ref{PShgSIGNSsimpson}).
Slices of the R-PDFL spectra (Fig. \ref{PShgSIGNSsimpson} C)) 
show a single splitting for the $\beta$-carbon with the order parameter value of 0.12,
and a superposition of a large and a very small splitting for the $\alpha$-carbon.
The larger splitting gives a order parameter value of 0.09, while the numerical value
from the small splitting cannot resolved with the available resolution.
Since only the absolute values of the PS headgroup order parameters were measured
previously \cite{browning80,roux91}, we used the S-DROSS experiment \cite{gross97} to determine the signs of
the order parameters. The S-DROSS slice for the $\beta$-carbon (Fig. \ref{PShgSIGNSsimpson} D)) clearly shows that the
order parameter is negative, which is confirmed by SIMPSON simulations.
The beginning of the S-DROSS slice suggests that the higher order parameter of
the $\alpha$-carbon is positive and the deviation towards negative values with the longer T$_1$ times suggests
that the smaller order parameter is negative. This is confirmed by a SIMPSON simulation
where the value of -0.02 was taken from $^2$H NMR experiment \cite{roux91} for the smaller order parameter.
The literature value was used because the
resolution of our experiment was not sufficient to determine the
small value of the order parameter. The S-DROSS curve from
SIMPSON simulation with a positive value for the smaller order parameter
(dashed grey in Fig. \ref{PShgSIGNSsimpson} D)) did not agree with the experiment, 
confirming the interpretation that the smaller order parameter is negative.
\begin{figure}[!htb]
  \centering
  \includegraphics[width=8.5cm]{../scratch/figIDEA.eps}
  \caption{\label{PShgSIGNSsimpson}
    (a) The headgroup region of the INEPT spectrum with headgroup and glycerol backbone carbons assigned.
    (b) 2D R-PDLF spectra for headgroup and glycerol backbone regions.
    (c) Slices for $\alpha$ and $\beta$ barbons.
    (d) Experimental SDROSS data (points) and SIMPSON simulations (lines).
    Order parameter values of -0.12 for the $\beta$-carbon, and 0.09 and -0.02
    for the larger and smaller $\alpha$-carbon slittings were used in the
    SIMPSON calculations. The S-DROSS curve from SIMPSON simulation with positive value
    for the smaller order parameter (dashed grey).
  }
  \todo{This is preliminary figure, should be polished.}
  \todo{Should we show slices for all the analyzed carbons in (c)?}
\end{figure}

@@

The headgroup and glycerol backbone order parameters of 
POPS measured in this work are in good agreement with the previously reported
values from $^2$H NMR experiments of DOPS \cite{browning80} (Fig. \ref{HGorderParameters}).
The $\beta$-carbon order parameter is significantly more negative and $\alpha$-carbon
experiences a significant forking in PS headgroup when compared with
the values previously measured for POPC \cite{ferreira13} (Fig. \ref{HGorderParameters}).  
These features have been intepreted to arise from a rigid PS headgroup
conformation, stabilized by hydrogen bonds or electrostatic
interactions \cite{browning80,buldt81}, but detailed structrural interpretation is not
available. 
%The detailed structural differences between the headgroups is, however, not known.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PCPScomp.pdf}
  \caption{\label{HGorderParameters}
    Headgroup and glycerol backbone order parameters of POPS measured in this work compared
    with values for DOPS ($^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
    POPC  ($^{13}$C NMR) \cite{ferreira13} from literature. Signs for PS order parameters
    as measured in this work and signs for PC as measured in Refs \cite{??,ferreira16}.
    %POPG from \cite{borle85} contains 10nM PIPES, DPPG from \cite{wohlgemuth80} contains  10mM PIPES and 100mM NaCl,
    %DPPE from \cite{seelig76}, E.coliPE and E.coliPG are from \cite{gally81}.
  }
  \todo{There should be values in \cite{roux90} which should be added.}
\end{figure}

@@

\todo{Also the discussion about POPS/OPPS issue with MacRog model should be added.}

@@

\subsection{Counterion binding to lipid bilayers containing PS lipids}
Membranes containing PS lipids are always accomppanied with counterions, which
modulate electrostatic interactions between lipids and other biomolecules. 
Counterions are also suggested screen the repulsion between charged lipid headgroups 
in MD simulations and reduce the area per lipid of PS bilayers to be smaller than in PC
bilayers~\cite{pandit02,mukhopadhyay04,pedersen06}. 
The counterion density profiles along membrane normal 
show significant differences between simulation models (Fig. \ref{NAdensPOPS}).
The strongest counterion binding, i.e., the lowest concentrations in bulk water,
are observed in MacRog, Berger and Lipid17/JC simulations.
CHARMM36, CHARMM36ua and Gromos-CKP models exhibit two local maxima in counterion
density, while a single maxima is observed in the other models. 
\todo{More detailed discussion may be possible after comparing monovalent ion binding
to bilayers between CHARMM simulations and experiments.}
Area per lipid is in agreement with experiments \cite{pan14} only
in the Gromos-CKP models, while other models give significantly lower values (Fig. \ref{NAdensPOPS}).
The difference cannot be explained by the electrostatic screening of the headgroup repulsion due to 
counterion binding because CHARMM36, CHARMM36ua and Slipid models give
smaller area per lipid than Gromos-CKP models with similar counterion binding affinity.
%The proposed trend is observed for Lipid17 model,
%for which the Joung-Cheatham ions give higher affinity and smaller area
%per molecule. However, the trend is not observed when comparing accross different
%force fields. For example, CHARMM36ua and MacRog give similar area per molecule
%but binding affinity is significantly higher in MacRog. 
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/NAdensPOPS.eps}
  \caption{\label{NAdensPOPS}
    Counterion densities of POPS lipid bilayer along the membrane normal from
    simulations with different force fields.
  }
\end{figure}

@@

The dehydrated complexes of PS headgroup and calcium ions can also lead to the
phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}.
Therefore, the Ca2+ binding affinity to PS lipids containing bilayers is easier
to study using mixtures diluted with PC lipids \cite{roux90,roux91}, where
the lipid-ion complexes and phase separation are not observed \cite{feigenson86,mattai89,roux90,roux91}.
The decrease of POPC headgroup order parameters as a function of Ca$^{2+}$ concentration in the
POPC:POPS (5:1) mixture is overestimated in all the tested simulation models
when compared with experiments \cite{roux90}, except in the CHARMM36 simulations with
special NBfix \cite{kim16} for calcium which underestimate the change (Fig. \ref{changesWITHCaClPS}).
According to the electrometer concept, this means that the calcium binding
seen in the ion density distributions along membrane normal (Fig. \ref{CAdensPCPSmixture})
is underestimated in the CHARMM36/NBfix model, but overestimated in the other tested models.
The overbinding of Ca$^{2+}$ in simulations is expected based on previous study of
PC lipid bilayers \cite{catte16}, but underestimated calcium binding affinity in
CHARMM36/NBfix model is surprising because CHARMM36 predicted overestimated
binding to PC bilayers. The difference can be explained by the NBfix interaction
parameters from Ref. \citenum{kim16}, incorporated in the parameters give by
the CHARMM-GUI at the time of running the simulations (January 2018).
These parameters underestimate also the binding of calcium to pure POPC bilayers
(Figs. \ref{OP_CHARMM_CaCl_POPC_NBFix} and \ref{density_profile_CHARMM_CaCl_POPC_NBFix}).

>>>>

Previous experimental studies have concluded that
PS headgroups are more rigid than phophocholines (PC)
due to the hydrogen bonding network or
electrostatic interactions \cite{browning80,buldt81}.
Multivalent cations and Li$^+$ are able to form strong
dehydrated molecular complexes with PS lipids,
while monovalent ions interact more weakly with PS
containing bilayers \cite{hauser77,kurland79,eisenberg79,hauser83,dluhy83,hauser85,feigenson86,mattai89,roux90,roux91,boettcher11}.
The dehydrated complexes of PS headgroup and calcium ions can also lead to the
phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}.
On the other hand, some studies propose that the specific binding
affinity is similar to the negatively charged and zwitterionic lipids and that
the increased cation binding to negatively charged lipid bilayer arise only due
to the increase of local cation concentration in the vicinity of membranes \cite{seelig90,sinn06}.
Dilution of bilayers with PC lipids makes PS headgroups
less rigid and reduces propensity for the formation of
strong complexes with multivalent ions \cite{browning80,buldt81,roux90,roux91}.
The molecular level interpretation of these observations is,
however, not available.

@@

}\label{PSsystems}

@@

}\label{mixedIONsystems}

@@

\subsection{Molecular dynamics simulations}
Molecular dynamics simulation data was collected using
the Open Collaboration method \cite{botan15}.
The NMRlipids project blog (\url{nmrlipids.blogspot.fi}) and
the GitHub repository (\url{github.com/NMRLipids/NMRlipidsIVotherHGs})
were used as the communication platforms.
The simulated systems are listed in 
Tables \ref{PSsystems} (pure PS systems without additional ions) 
and \ref{mixedIONsystems} (mixed PC:PS systems with various ions concentrations).
Simulation details are given in the SI.
The simulation data is also indexed in the
searchable database (\url{nmrlipids.fi}),
and in the NMRlipids/MATCH GitHub repository (\url{https://github.com/NMRLipids/MATCH}).

@@

The order parameters of $\alpha$ and $\beta$ carbons in PC lipids
can be used to measure the ion binding affinity because they 
decrease proportionally to the amound of bound positive
charge to a bilayer \cite{akutsu81,altenbach84,seelig87}.
This molecular electrometer concept is especially useful for 
the comparison between simulations and experiments because
the headgroup order parameters can be directly calculated from 
simulations~\cite{catte16}. Also the headgroup order parameters
of negatively charged PS and PG lipids exhibit systemic, but less
characterized dependence on the bound charge~\cite{borle85,macdonald87,roux86,roux90}.
Therefore, the ion binding affinity to negatively charged bilayers
can be better characterized by measuring the PC headgroup order parameters from 
mixed bilayers~\cite{roux86,roux90,roux91}, see section \ref{electrometerFORmixtures} in the supplementary information.

@@

\subsection{Headgroup and glycerol backbone order parameters of POPS from $^{13}$C NMR}

The INEPT and 2D R-PDLF experiments from POPS sample give well resolved spectras for all the
carbons in headgroup and glycerol backbone region, except for g$_3$ for which the resolution
was not sufficient to determine the numerical value of the order paramater (Fig. \ref{PShgSIGNSsimpson}).
Slices of the R-PDFL spectra (Fig. \ref{PShgSIGNSsimpson} C)) 
show a single splitting for the $\beta$-carbon with the order parameter value of 0.12,
and a superposition of a large and a very small splitting for the $\alpha$-carbon.
The larger splitting gives a order parameter value of 0.09, while the numerical value
from the small splitting cannot resolved with the available resolution.
Since the R-PDFL and previous $^2$H NMR experiments \cite{browning80,roux91} give 
only the absolute values of order parameters, we determined the signs of PS headgroup
and glycerol backbone order parameters using the S-DROSS experiment \cite{gross97}.
The S-DROSS slice for the $\beta$-carbon (Fig. \ref{PShgSIGNSsimpson} D)) clearly shows that the
order parameter is negative, which is confirmed by SIMPSON simulations.
The beginning of the S-DROSS slice suggests that the higher order parameter of
the $\alpha$-carbon is positive and the deviation towards negative values with the longer T$_1$ times suggests
that the smaller order parameter is negative. This is confirmed by a SIMPSON simulation
where the value of -0.02 was taken from $^2$H NMR experiment \cite{roux91} for the smaller order parameter.
The literature value was used because the
resolution of our experiment was not sufficient to determine the
small value of the order parameter. The S-DROSS curve from
SIMPSON simulation with a positive value for the smaller order parameter
(dashed grey in Fig. \ref{PShgSIGNSsimpson} D)) did not agree with the experiment, 
confirming the interpretation that the smaller order parameter is negative.
\begin{figure}[!htb]
  \centering
  \includegraphics[width=8.5cm]{../scratch/figIDEA.eps}
  \caption{\label{PShgSIGNSsimpson}
    (a) The headgroup region of the INEPT spectrum with headgroup and glycerol backbone carbons assigned.
    (b) 2D R-PDLF spectra for headgroup and glycerol backbone regions.
    (c) Slices for $\alpha$ and $\beta$ barbons.
    (d) Experimental SDROSS data (points) and SIMPSON simulations (lines).
    Order parameter values of -0.12 for the $\beta$-carbon, and 0.09 and -0.02
    for the larger and smaller $\alpha$-carbon slittings were used in the
    SIMPSON calculations. The S-DROSS curve from SIMPSON simulation with positive value
    for the smaller order parameter (dashed grey).
  }
  \todo{This is preliminary figure, should be polished.} \\
  \todo{Should we show slices for all the analyzed carbons in (c)?}
\end{figure}

@@

The headgroup and glycerol backbone order parameters of 
POPS measured in this work are in good agreement with the previously reported
values from $^2$H NMR experiments of DOPS \cite{browning80} (Fig. \ref{HGorderParameters}).
The $\beta$-carbon order parameter is significantly more negative and $\alpha$-carbon
experiences a significant forking in PS headgroup when compared with
the values previously measured for POPC \cite{ferreira13} (Fig. \ref{HGorderParameters}).  
These features have been intepreted to arise from a rigid PS headgroup
conformation, stabilized by hydrogen bonds or electrostatic
interactions \cite{browning80,buldt81}, but detailed structrural interpretation is not
available. 
%The detailed structural differences between the headgroups is, however, not known.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PCPScomp.pdf}
  \caption{\label{HGorderParameters}
    Headgroup and glycerol backbone order parameters of POPS measured in this work compared
    with the values from DOPS ($^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
    POPC  ($^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
    are measured in this work. Signs of the PC order parameters are measured in Ref.~\cite{ferreira16}.
  }
\end{figure}

@@

\todo{POPS/OPPS issue with MacRog model is in progress: https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/16}

@@

\subsection{Counterion binding to lipid bilayers containing PS lipids}
Membranes containing PS lipids are always accomppanied with counterions, which
modulate electrostatic interactions between lipids and other biomolecules. 
Counterions are also suggested screen the repulsion between charged lipid headgroups 
in MD simulations and reduce the area per lipid of PS bilayers to be smaller than in PC
bilayers~\cite{pandit02,mukhopadhyay04,pedersen06}. 
The counterion density profiles along membrane normal 
show significant differences between simulation models (Fig. \ref{NAdensPOPS}).
The strongest counterion binding, i.e., the lowest concentrations in bulk water,
are observed in MacRog, Berger and Lipid17/JC simulations.
CHARMM36, CHARMM36ua and Gromos-CKP models exhibit two local maxima in counterion
density, while a single maxima is observed in the other models. 
\todo{More detailed discussion may be possible after comparing monovalent ion binding
to bilayers between CHARMM simulations and experiments.
Also, section S6 should be finished.}
Area per lipid is in agreement with experiments \cite{pan14} only
in the Gromos-CKP models, while other models give significantly lower values (Fig. \ref{NAdensPOPS}).
The difference cannot be explained by the electrostatic screening of the headgroup repulsion due to 
counterion binding because CHARMM36, CHARMM36ua and Slipid models give
smaller area per lipid than Gromos-CKP models with similar counterion binding affinity.
%The proposed trend is observed for Lipid17 model,
%for which the Joung-Cheatham ions give higher affinity and smaller area
%per molecule. However, the trend is not observed when comparing accross different
%force fields. For example, CHARMM36ua and MacRog give similar area per molecule
%but binding affinity is significantly higher in MacRog. 
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/NAdensPOPS.eps}
  \caption{\label{NAdensPOPS}
    Counterion densities of POPS lipid bilayer along the membrane normal from
    simulations with different force fields.
  }
\end{figure}

@@

Ion binding affinity to PS containing membranes can be most conveniently measured 
from PC:PS lipid mixtures where the lipid-ion complexes and phase separation are 
not observed \cite{feigenson86,mattai89,roux90,roux91}. Also, the headgroup order 
parameters of PC can be used to detect ion binding affinity to the mixed lipid 
bilayers, see section~\ref{electrometerFORmixtures}. As expected from the previous study of PC lipid bilayers \cite{catte16}, 
the decrease of POPC headgroup order parameters in POPC:POPS mixtures as a function of Ca$^{2+}$ concentration 
is overestimated with respect to experiments \cite{roux90} in almost all the 
tested simulation models (Fig. \ref{changesWITHCaClPS}), indicating overestimated calcium binding
binding affinity. Only exception is the CHARMM36 model with the special NBfix~\cite{kim16} interaction, 
incorporated in the parameters give by the CHARMM-GUI at the time of running the simulations (January 2018),
which underestimates the calcium binding affinity. With this interaction, the 
calcium and sodium binding affinities to POPC bilayer are equally weak (see section S7), in contrast to experimental 
data~\cite{cevc90,akutsu81,altenbach84}. Therefore, we conclude that the ion binding peaks in
density distributions along membrane normal (Fig. \ref{CAdensPCPSmixture}) are underestimated 
in the CHARMM36/NBfix model, but overestimated in the other tested models.

### 2018-08-09 Manuscript/manuscriptPSsuppl.tex
Updates in the text

The electrometer concept is based on the empirical observations that
the order parameters of $\alpha$ and $\beta$ carbons in PC lipid headgroup
decrease (increase) proportionally to the bound positive (negative) 
charge~\cite{akutsu81,altenbach84,seelig87,scherer89}. 
The concept is also valid for lipid mixtures where PC order parameters 
increase when mixed with anionic lipids, in contrast to the mixtures with neutral or
cationic surfactants~\cite{seelig87, scherer87} (Fig. \ref{HGorderparametersPCvsPEPSPGchol}).
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPEPSPGchol.eps}
  \caption{\label{HGorderparametersPCvsPEPSPGchol}
    PC headgroup order parameters from experiments of mixtures with
    PE, PS, PG and cholesterol \cite{scherer87,scherer89,ferreira13}.
    Signs are determined as discussed in \cite{botan15,ollila16}.
  }
\end{figure}

Based on the electrometer concept, the headgroup order parameters of
PC lipids can be used to measure the ion binding affinity to lipid 
bilayers \cite{akutsu81,altenbach84,seelig87}. Changes of the headgroup order 
parameters of negatively charged PS and PG lipids with bound charge are also systematic,
but less well characterized~\cite{borle85,macdonald87,roux86,roux90}.
Therefore, the ion binding affinity to negatively charged bilayers
can be better characterized measuring the PC headgroup order parameters from 
mixed bilayers. The PC headgroup order parameters in mixtures with anionic
lipids are larger than in pure PC lipid because addition of
negative charges increase the order parameters (Figs. \ref{HGorderparametersPCvsPEPSPGchol} 
and \ref{OrderParametersWithCaCl}). Upon addition of CaCl$_2$, the PC headgroup order 
parameters in the mixed bilayers decrease and reach the values of pure PC bilayer close 
to the CaCl$_2$ concentrations of $\sim$ 50-300mM, depending on the amount of negatively charged
lipids in the mixture (Fig. \ref{OrderParametersWithCaCl}). At this point the charge of
bound Ca$^{2+}$ probably cancels the charge of anionic negative lipids and the bilayer 
has neutral charge. The overcharging then occurs with larger concenterations.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/LIPIDSwithCaCl.eps}
  \includegraphics[width=8.0cm]{../Figs/LIPIDSwithCaClBELOW1M.eps}
  \caption{\label{OrderParametersWithCaCl}
    PC headgroup order parameters as a function of CaCl concentration from experiments 
    different mole fractions of negatively charged lipids.
    Pure DPPC data from \cite{akutsu81}, pure POPC data from \cite{altenbach84}, 
    POPC:POPS mixture data from \cite{roux90}, POPC:POPG mixture data with 0.1M NaCl from \cite{macdonald87}
    and POPC:POPG mixture data without NaCl from \cite{borle85}.
    The right column shows the results zoomed to concentrations below 1M.
  }
\end{figure}
As expected, the more pronounced order parameter decrease is observed in systems with more
negatively charged lipids (Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}) indicating 
higher binding affitiny of cations into the negatively charged membranes. In conclusion,
the consistency of the experimental results suggests that the electrometer concept can
be used to determine the cation binding also to the lipid bilayers containing negatively
charged lipids.  
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaCl.eps}
  \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}
    The change of PC headgroup order parameters as a function of CaCl$_2$
    measured from bilayers containing different amount of negatively charged lipids.
    The values are taken from 2H NMR experiments reported in the
    literature (DPPC \cite{akutsu81}, POPC \cite{altenbach84}, POPC:POPS (5:1) \cite{roux90},
    POPC:POPG  mixtures with 0.1M NaCl \cite{macdonald87}
    and POPC:POPG (1:1) without NaCl \cite{borle85}).
    As expected, the decrease of order parameters with the added CaCl$_2$
    is more pronounced for systems with larger fraction of
    negatively charged lipids, indicating larger amount amount
    of bound cations.
  }
\end{figure}

\pagebreak
\section{Calibration of PC headgroup order parameter response to the bound cations}\label{electrometerCALIBRATION}
Before using the headgroup order parameters to compare ion binding affinity between simulations
and experiments, it is important to quantify the response of the order parameters to the
bound charge in simulations~\cite{catte16,melcr18}. Comparison of the headgroup order parameter
response to the cationic surfactants in POPC bilayer to the experiments \cite{scherer89} shows
that the both order parameters are too sensitive to the bound charge in the Lipid14 model,
while CHARMM36 gives better agreement for the $\alpha$ carbon (Fig. \ref{CHANGESwithDHMDMAB}).
The ratio $\Delta S_{\rm CH}^\beta / \Delta S_{\rm CH}^\alpha$ from Lipid14 model was in good agreement 
with experiments in the previous study~\cite{catte16} because both order parameters are equally too 
sensitive to the bound cations (Fig. \ref{CHANGESwithDHMDMAB}). The ratio was, however, overestimated
for the CHARMM36 model because the $\beta$ order parameter relatively more sensitive than the $\alpha$
order parameter. These have to be taken into account when analysing the ion binding affinities
using headgroup order parameters in simulations.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGopsDHMDMAB.eps}
  \caption{\label{CHANGESwithDHMDMAB}
  The response of headgroup order parameters to the fixed amount of cationic surfactants in
  POPC bilayer is compared between simulations and experiments \cite{scherer89}.}
\end{figure}

@@

>>>>

The electrometer concept is based on the empirical observations that
the order parameters of $\alpha$ and $\beta$ carbons in PC lipid headgroup
decrease (increase) proportionally to the bound positive (negative) 
charge~\cite{akutsu81,altenbach84,seelig87,scherer89} (Fig. \ref{HGorderparametersPCvsPEPSPGchol}). 
Therefore, the headgroup order parameters can be used to measure the
ion binding affinity to lipid bilayers containing PC
lipids~\cite{akutsu81,altenbach84,borle85,seelig87,macdonald87,roux90}.
Changes of the headgroup order parameters of other lipids with bound charge,
like negatively charged PS and PG lipids,  are also systematic,
but less well characterized~\cite{borle85,macdonald87,roux86,roux90}.
Therefore, the ion binding affinity to negatively charged bilayers
can be better characterized measuring the PC headgroup order parameters from 
mixed bilayers~\cite{borle85,roux86,macdonald87,roux90,roux91}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPEPSPGchol.eps}
  \caption{\label{HGorderparametersPCvsPEPSPGchol}
    Headgroup order parameters of POPC measured from mixtures with
    PS (bovine brain), POPG, POPE, cholesterol and cationic surfactant~\cite{scherer87,scherer89,ferreira13}.
    Signs are taken from separate experiments~\cite{ollila16,ferreira16}.
  }
\end{figure}

When comparing ion binding measured from the PC headgroup order parameters between
neutral and anionic membranes, it is important to note that the order parameters
increase due to the addition of negative charged lipids according to the electrometer
concept~\cite{seelig87, scherer87} (Fig. \ref{HGorderparametersPCvsPEPSPGchol}).
Therefore, the headgroup order parameters of POPC in mixtures with anionic
lipids are larger than in pure POPC lipid bilayer, which is also seen in the ion binding
experiments from different mixtures in the absence of added calcium (Fig. \ref{OrderParametersWithCaCl}).
Upon addition of CaCl$_2$, the order parameters decrease and reach the values of pure PC bilayer close 
to the CaCl$_2$ concentrations of $\sim$ 50-300mM, depending on the amount of negatively charged
lipids in the mixture (Fig. \ref{OrderParametersWithCaCl}).
Around these concentrations the positive charge of bound Ca$^{2+}$ cancels
the negative charge lipids, resulting to a neutral membrane. 
Above such concentrations, the specific binding of calcium leads to
the overcharging of the membrane.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/LIPIDSwithCaCl.eps}
  \includegraphics[width=8.0cm]{../Figs/LIPIDSwithCaClBELOW1M.eps}
  \caption{\label{OrderParametersWithCaCl}
    Headgroup order parameters of POPC as a function of CaCl concentration from experiments 
    different mole fractions of negatively charged lipids (left column).
    Right column shows the same data zoomed to the concentrations below 1M.
    Data for Pure DPPC from~\cite{akutsu81},
    for pure POPC from~\cite{altenbach84}, 
    for POPC:POPS (5:1) mixture from~\cite{roux90},
    for POPC:POPG (8:2,1:1) mixtures with 0.1M NaCl from~\cite{macdonald87}
    and for POPC:POPG (1:1) mixture data without NaCl from \cite{borle85}.
  }
\end{figure}

Because the POPC headgroup order parameters in mixtures with different amounts of anionic lipids
but without additional salt are not equal, the binding affinity of calcium can be better compared
by plotting the changes of order parameters as a function of added calcium.
As expected, such a plot reveals more pronounced order parameter decrease in systems
with larger fractions of negatively charged lipids (Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}),
indicating an increase in the calcium binding affinity with the increasing amount of negatively charged
lipids in membranes. In conclusion, the presented empirical comparison of headgroup order parameter changes
from various mixtures of POPC and anionic lipids with added calcium gives physically
consistent results, suggesting that the electrometer concept can be used to determine
the cation binding affinity also to the lipid bilayers containing mixtures of PC and anionic lipids.  
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaCl.eps}
  \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}
    Changes of POPC headgroup order parameters as a function of CaCl$_2$
    measured from mixed bilayers containing different amounts of anionic lipids.
    The original data is the same as in figure~\ref{OrderParametersWithCaCl}.
%    The values are taken from 2H NMR experiments reported in the
%    literature (DPPC \cite{akutsu81}, POPC \cite{altenbach84}, POPC:POPS (5:1) \cite{roux90},
%    POPC:POPG  mixtures with 0.1M NaCl \cite{macdonald87}
%    and POPC:POPG (1:1) without NaCl \cite{borle85}).
  }
\end{figure}

\pagebreak
\section{Calibration of PC headgroup order parameter response to the bound cations}\label{electrometerCALIBRATION}
Before using the headgroup order parameters to compare ion binding affinity between simulations
and experiments, the response of the order parameters to the bound charge in simulations should be quantified
against experiments~\cite{catte16,melcr18}. In our previous work~\cite{catte16},
the ratio $\Delta S_{\rm CH}^\beta / \Delta S_{\rm CH}^\alpha$ was in good agreement with the experiments~\cite{akutsu81}
in the Lipid14 model, but was underestimated by other models. In the more recent study~\cite{melcr18},
the headgroup order parameter responses were compared more carefully with the experiments of cationic
surfactants in POPC bilayer~\cite{scherer89}. The comparison reveals that the both headgroup
order parameters in the Lipid14 model are too sensitive to the bound charge,
while CHARMM36 gives better agreement for the $\alpha$ carbon (Fig. \ref{CHANGESwithDHMDMAB}).
Therefore, the headgroup order parameter response to the bound charge is actually more realistic
in CHARMM36 model than in the Lipid14, where both order parameters are equally too 
sensitive to the bound cations giving a good result for the $\Delta S_{\rm CH}^\beta / \Delta S_{\rm CH}^\alpha$ ratio.
The ratio was overestimated for the CHARMM36 model because the $\beta$-carbon order parameter
is relatively more sensitive than the $\alpha$-carbon order parameter. These results have to be taken
into account when analysing the ion binding affinities using headgroup order parameters in simulations.
However, we conclude that the discrepancies arising from the sentivity of lipid headgroup to
bound charge are typically smaller than then discrepancies arising from ion binding affinity.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGopsDHMDMAB.eps}
  \caption{\label{CHANGESwithDHMDMAB}
  Responses of headgroup order parameters to the fixed amount of cationic surfactants in
  POPC bilayer from simulations and experiments \cite{scherer89}.
  The simulation results for Lipid14 are directly from Ref.~\citenum{melcr18}.
  The CHARMM36 simulation data and details are available from Ref.~\citenum{CHARMM36cationicSURF}.
}
\end{figure}

@@

\todo{This is to be removed once the new data is available:
https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/16}
